OneD: increasing reproducibility of Hi-C Samples with abnormal karyotypes

### 2017-11-24 manuscript_nar.tex
Last edits up to results (still need to do last sections).

Several methods have been proposed to remove biases in Hi-C experiments
\citep{schmitt2016genome}. The first strategy is to model biases
explicitly from a defined set of local genomic features, such as the G+C
content. This approach is used in the method of
\cite{yaffe2011probabilistic} and in Hicnorm by \cite{hu2012hicnorm}. The
second strategy is to implicitly correct unknown biases by enforcing some
regularity condition on the data. This approach is used in the
Iterative Correction and Eigenvector decomposition method (\textit{ICE})
of \cite{imakaev2012iterative}, whereby the total amount of contacts of
every bin is imposed to be the same. \textit{ICE} is currently the most
popular method, due in part to its speed.

Neither of these strategies were designed for cell types with karyotypic
aberrations, most common in cancer. Yet, Hi-C is very sensitive to
aneuploidy, copy number variations and translocations.  Actually, these
aberrations have so much influence on the outcome that they can be used as
signature to re-assemble the target genome \citep{korbel2013genome}. An
additional complication is that karyotypic aberrations are not
experimental biases, so it is unclear whether they should be corrected at
all or be considered part of the biological signal.

@@

So far, the only attempt to address the issue was the chromosome-adjusted
Iterative Correction Bias method (\textit{caICB}) of
\cite{wu2016computational}. However, \textit{caICB} applies a uniform
chromosome-wide copy number correction, effectively excluding the numerous
cases of partial aneuploidy and regional copy number variations.

@@

Our approach is to model the tally of contacts for each bin, thus reducing
the matrix to a one dimension score (hence the name \textit{OneD}). We
assume that the total number of contacts per bin ($t_{i}$) can be
approximated by a negative binomial distribution. This choice is sensible
because the amount of contacts is a discrete variable and because the
negative binomial distribution allows for overdispersion. We further
assume that the explicit sources of bias have independent contributions to
the mean of the distribution for a given bin ($\lambda_i$).

Given that this relationship might not be linear (see for instance
Figure~\ref{fig:totals}A), we allowed a smooth representation
using thin plate penalized regression splines \citep{wood2003thin} in a
generalized additive model \citep{wood2011fast}. The model can be
parametrized as

\begin{align*}
t_i = \sum_j^n{x_{ij}} &\sim  NB(\lambda_i, \theta) \\
\log(\lambda_i) &\propto \sum_{k}{f_k(z_k)}
\end{align*}

\noindent
where $x_{ij}$ is the raw number of contacts between bins $i$ and $j$, and
$z_k$ is the additive bias of genomic feature $k$. The smooth functions
$\{f_k(\cdot)\}$ are estimated jointly with the negative binomial
dispersion parameter $\theta$ using the the default arguments of the
\texttt{mgcv::s} and \texttt{mgcv::nb} functions \citep{wood2011fast} of the R
software \citep{coreteam2014r}.

Once the parameters of the model are determined, the estimated means
$\{\lambda_i\}$ are rescaled to obtain a correction vector
$\{\lambda_i'\}$ that can be used to compute the corrected counts
($\hat{x}_{ij}$).

\begin{align}
\notag
\lambda_i' &= \frac{\lambda_i}{\sum_j^n{\lambda_j}/n} \\
\label{eq:xhat}
\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}
\end{align}

@@

Briefly, a hidden Markov model with emissions distributed as a Student's
$t$ variable is fitted on the corrected total amount of contacts per bin
($\hat{t}_i = \sum_j^n{\hat{x}_{ij}}$). The model consists of 8 states that correspond to 1, 2, 3,
4, 5, 6, 7, and `8 or more' copies of the target, for a total of 4
emission parameters (a single position parameter for states 1-7, a
position parameter for state `8 or more', a single standard deviation for
all the states and a single degree of freedom for all the states) and 56
transition parameters.

The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
convergence, following a previously described implementation
\citep{filion2010systematic}.  The Viterbi path is then computed and
corresponds to the inferred copy number of each bin ($c_i$).

A correction equal to the square root of the copy number is then applied
to the whole matrix. More specifically, each cell is updated to

\begin{equation}
\label{eq:cnvnorm}
\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}.
\end{equation}

@@

To test the correction of biases, we gathered a set of
published \citep[][BioRxiv: https://doi.org/10.1101/132456]{ledily2014distinct, encode2012integrated, rao20143d,
lin2012global, dixon2012topological} and
unpublished Hi-C data of different cell types and organisms
(Supplementary Table~1).

We used several experiments comprising T47D breast cancer cell lines (6
samples), K562 leukemia cell lines (8 samples), both with aberrant
karyotypes; mouse primary B cells (6 samples) and ES cells (7
samples), both with normal diploid karyotypes; GM12878 B-lymphoblastoid cells
(58 samples), with diploid karyotype; BT474, MCF10, MCF-7 and SKBR3 breast cell
lines (1 sample each). The experiments were
carried out in different laboratories, following either the original Hi-C
protocol \citep{lieberman2009comprehensive} or the newer \textit{in situ}
version \citep{rao20143d}, and using different restriction enzymes (DpnII,
HindIII, MspI, MboI and NcoI). We also used array-based copy-number
segmentation of the two cell lines obtained from the COSMIC database
\citep{forbes2010cosmic}.

%\begin{landscape}
%
%\begin{table*}
%\label{tab:samples}
%{\begin{tabular}{ccccccc}
%  \hline
%  \textbf{Sample ID} & \textbf{Cell type} & \textbf{Application} &
%  \textbf{RE} & \textbf{Sequencing core} & \textbf{Set(s)} &
%  \textbf{Source} \\
%  \hline
%dc3a1e069\_51720e9cf & T47D & \textit{in situ} Hi-C &
%  DpnII & CRG & T47D, K562 & NA \\
%b1913e6c1\_51720e9cf & T47D & \textit{in situ} Hi-C &
%  DpnII & CRG & T47D & NA \\
%dc3a1e069\_ec92aa0bb & T47D & \textit{in situ} Hi-C &
%  DpnII & CRG & T47D, K562 & NA \\
%HindIII\_T0 & T47D & dilution Hi-C &
%  HindIII & CRG & T47D & SRR1054341 \\
%NcoI\_T0    & T47D & dilution Hi-C &
%  NcoI      & CRG & T47D & SRR1054343 \\
%ENCLB758KFU & T47D & dilution Hi-C &
%  HindIII   & UMass & T47D & ENCLB758KFU \\
%ENCLB183QHG & T47D & dilution Hi-C &
%  HindIII   & UMass & T47D & ENCLB183QHG \\
%HIC069  & K562 & \textit{in situ} Hi-C &
%  MboI & Baylor & T47D, K562 &    SRR1658693 \\
%HIC070  & K562 & \textit{in situ} Hi-C &
%  MboI & Baylor & T47D, K562 &    SRR1658694 \\
%HIC071  & K562 & \textit{in situ} Hi-C &
%  MboI & Baylor & K562 & SRR1658695,SRR1658696 \\
%HIC074  & K562 & \textit{in situ} Hi-C &
%  MboI & Baylor & K562 & SRR1658701,SRR1658702 \\
%b7fa2d8db\_bfac48760 & B cell  & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%fc3e8b36a\_7bf1bf374 & ES cell & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%b7fa2d8db\_7284b867a & B cell  & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%fc3e8b36a\_38bfd1b33 & ES cell & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%b7fa2d8db\_58e812fc2 & B cell  & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%fc3e8b36a\_c990a254e & ES cell & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%b7fa2d8db\_73f11d923 & B cell  & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%fc3e8b36a\_4bf044f18 & ES cell & \textit{in situ} Hi-C &
%  DpnII & CRG & mm10 & GSE96611 \\
%GSM987817 & B cell  & dilution Hi-C &
%  HindIII & UCSD & mm10 & SRR543428-SRR543431 \\
%GSM987818 & B cell  & dilution Hi-C &
%  HindIII & UCSD & mm10 & SRR543432-SRR543442 \\
%GSM862720 & ES cell & dilution Hi-C &
%  HindIII & LICR & mm10 & SRR443883-SRR443885 \\
%GSM862721 & ES cell & dilution Hi-C &
%  HindIII & LICR & mm10 & SRR400251-SRR400256 \\
%GSM862722 & ES cell & dilution Hi-C &
%  NcoI    & LICR & mm10 & SRR443886-SRR443888 \\
%  \hline
%\end{tabular}}{}
%\caption{Hi-C data sets used in this study. Sample ID: Unique sample
%identifier in the respective projects. Cell type: source of the biological
%sample. Application: Hi-C protocol. RE: Restriction enzyme used for the
%digestion. Sequencing core: Laboratory performing the experiment (CRG:
%Centre for Genomic Regulation; UMass: University of Massachusetts; Baylor:
%Baylor College of Medicine; UCSD: University of California San Diego;
%LICR: Ludwig Institue for Cancer Research). Set(s): Sets in which the
%sample was included (see text for detail).  Source: SRA, GEO or ENCODE raw
%data identifier.}
%\end{table*}
%
%\end{landscape}

%\subsection{Data processing}

All data were processed through a pipeline based on TADbit
\citep{serra2016structural}. Briefly, after controlling the quality of
FASTQ files, paired-end reads were mapped to the corresponding reference
genome (hg38 and mm10) taking into account the restriction enzyme site.
Non-informative contacts were removed applying the following TADbit
filters: \texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
\texttt{extra-dangling-ends}, \texttt{duplicated} and
\texttt{random-breaks}. For more details, see the methods section of
(BioRxiv: https://doi.org/10.1101/132456). In addition, the pipeline is
available from the Supplementary Material published by
\cite{quilez2017managing}.

We developed the routines contained in the \texttt{dryhic} R package
(available at http://www.github.com/qenvio/dryhic) to efficiently create
sparse representations of contact matrices and further apply
\textit{vanilla}, \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. All the results are based on a
resolution of 100~Kbp, but we found no major differences at different
resolutions (not shown).

@@

Simulations were based on altering four experimental Hi-C samples obtained
from diploid mouse cell lines (b7fa2d8db\_bfac48760, fc3e8b36a\_7bf1bf374,
GSM987817 and GSM862720), either B cell or ES cell, and from either
CRG or UCSC. The simulation strategy was the following: First we selected
uniformly at random the amount of copy number break points (from 3 to 10).
We then placed the break points along chromosomes 18 and 19 uniformly at
random. Next, we assigned a copy number (2, 3, 4 or 10 with equal
probability) to each segment delimited by the break points. We computed
the outer product of this simulated copy number profile and multiplied it
element-wise with the original contact matrices of chromosomes 18 and 19.
The resulting matrices were used as input for correction methods. For each
simulation (100 in total), we measured the pairwise reproducibility score
defined by \citet{yan2017hicspector} and computed the average for pairs
from the same cell type minus the average for pairs from different cell
types.

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mbp in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined by
\cite{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
\citet{yan2017hicspector} sums the distances between the leading twenty
eigenvectors of the Laplacian of the Hi-C matrix. This approach borrows
the concepts of spectral clustering \citep{von2007tutorial} and amounts to
comparing high level features of the matrix.

We defined five data sets to measure experimental reproducibility after
normalization: The first contained the samples from T47D plus two samples
from K562, the second contained the samples from K562 plus two samples
from T47D, the third contained all the mouse samples, the fourth contained all
GM12878 samples, the fifth contained one sample of each breast cells (see Supplementary
Table~1 for details). Given a set of experiments and a metric, we first
computed all pairwise combinations between experiments and then classified
the comparisons according to the characteristics of each pair (cell type,
protocol, batch and treatment).

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. To test the ability of the different methods to separate samples
from different cell origin we generated receiver operating characteristic (ROC) curves
using the similarity metric (e.g. reproducibility score) as the classifier score
and the relative cell type (i.e. same or different cell type) as the binary
classification. ROC curves were computed using the \texttt{ROCR} package
\citep{sing2005rocr} for the three first sets (those including samples of
different cell types).

>>>>

Several methods have been proposed to remove biases in Hi-C experiments
\citep{schmitt2016genome}. The first strategy is to model biases
explicitly from a defined set of local genomic features, such as the G+C
content. This approach is used in the method of
\citet{yaffe2011probabilistic} and in HiCNorm by \citet{hu2012hicnorm}.
The second strategy is to implicitly correct unknown biases by enforcing
some regularity condition on the data. This approach is used in the
Iterative Correction and Eigenvector decomposition method (\textit{ICE})
of \citet{imakaev2012iterative}, whereby the total amount of contacts of
every bin is imposed to be the same. \textit{ICE} is currently the most
popular method, due in part to its speed.

Neither of these strategies were designed for cell types with karyotypic
aberrations, most common in cancer. Yet, Hi-C is very sensitive to
aneuploidy, copy number variations and translocations. Actually, these
aberrations have so much influence on the outcome that they can be used as
signatures to re-assemble the target genome \citep{korbel2013genome}. An
additional complication is that karyotypic aberrations are not
experimental biases, so it is unclear whether they should be corrected at
all or be considered part of the biological signal.

@@

So far, the only attempt to address the issue was the chromosome-adjusted
Iterative Correction Bias method (\textit{caICB}) of
\citet{wu2016computational}. However, \textit{caICB} applies a uniform
chromosome-wide copy number correction, effectively excluding the numerous
cases of partial aneuploidy and regional copy number variations.

@@

Our approach is to model the total number of contacts for each bin, thus
reducing the matrix to a one-dimension score (hence the name
\textit{OneD}) referred to as the `contact profile'. We assume that the
total number of contacts per bin $t_{i}$ can be approximated by a
negative binomial distribution. This choice is sensible because the amount
of contacts is a discrete variable and because the negative binomial
distribution allows for overdispersion. We further assume that the
explicit sources of bias have independent contributions to the mean of the
distribution for a given bin $\lambda_i$.

Given that this relationship might not be linear (see for instance
Figure~\ref{fig:totals}A), we allowed a smooth representation using thin
plate penalized regression splines \citep{wood2003thin} in a generalized
additive model \citep{wood2011fast}. Thin plate regression splines are
`isotropic' because rotation of the covariate co-ordinate system does not
change the result of smoothing and `low rank' because they have far fewer
coefficients than there are data points to smooth. The model can be
parametrized as
\begin{gather*}
t_i = \sum_{j=1}^n{x_{ij}} \sim  NB(\lambda_i, \theta) \text{ and} \\
\log(\lambda_i) \propto \sum_{k}{f_k(z_{k,i})},
\end{gather*}
where $x_{ij}$ is the raw number of contacts between bins $i$ and $j$, and
$z_{k,i}$ is the additive bias of genomic feature $k$ in bin $i$.

The smooth functions $\{f_k(\cdot)\}$ are estimated through the thin plate
spline penalty, jointly with the negative binomial dispersion parameter
$\theta$ using the the default arguments of the \texttt{mgcv::s} and
\texttt{mgcv::nb} functions \citep{wood2011fast} of the R software
\citep{coreteam2014r}.

Once the parameters of the model are determined, the estimated means
$\{\lambda_i\}$ are rescaled to obtain a correction vector
$\{\lambda_i'\}$ that can be used to compute the corrected counts
$\hat{x}_{ij}$.
\begin{align}
\notag
\lambda_i' &= \frac{\lambda_i}{\sum_j^n{\lambda_j}/n} \\
\label{eq:xhat}
\hat{x}_{ij} &= \frac{x_{ij}}{\sqrt{\lambda_i'\lambda_j'}}
\end{align}

@@

Briefly, a hidden Markov model with emissions distributed as a Student's
$t$ variable is fitted on the corrected total amount of contacts per bin
$\hat{t}_i = \sum_{j=1}^n{\hat{x}_{ij}}$. The model consists of 8
states that correspond to 1, 2, 3, 4, 5, 6, 7, and `8 or more' copies of
the target bin, for a total of 4 emission parameters (a single position
parameter for states 1-7, a position parameter for state `8 or more', a
single standard deviation for all the states and a single degree of
freedom for all the states) and 56 transition parameters.

The model is fitted with the Baum-Welch algorithm \citep{baum1966} until
convergence, following a previously described implementation
\citep{filion2010systematic}.  The Viterbi path is then computed and
corresponds to the inferred copy number of each bin $c_i$.

A correction equal to the square root of the copy number is then applied
to the whole matrix. More specifically, the entry at position $(i,j)$ is
updated to
\begin{equation}
\label{eq:cnvnorm}
\hat{x}_{ij}^* = \frac{\hat{x}_{ij}}{\sqrt{c_ic_j}}.
\end{equation}

@@

We gathered a set of published Hi-C data \citep[][BioRxiv:
https://doi.org/10.1101/132456]{ledily2014distinct, encode2012integrated,
rao20143d, lin2012global, dixon2012topological} and unpublished data of
different cell types and organisms (Supplementary Table~1).

The Hi-C experiments were performed in T47D breast cancer cell lines
(aberrant karyotype, six samples), K562 leukemia cell lines (aberrant
karyotype, eight samples), mouse primary B cells (normal karyotyp, six
samples) ES cells (normal karotype, seven samples), GM12878
B-lymphoblastoid cells (normal karyotype, fifty eight samples), BT474
breast cancer cell line (aberrant karyotype, one sample), MCF10 breast
cancer cell line (aberrant karyotype, one sample), MCF-7 cancer cell line
(aberrant karyotype, one sample) and SKBR3 breast cell lines (aberrant
karyotype, one sample).

The experiments were carried out in different laboratories, following
either the original Hi-C protocol \citep{lieberman2009comprehensive} or
the newer \textit{in situ} version \citep{rao20143d}, and using different
restriction enzymes (DpnII, HindIII, MspI, MboI and NcoI). In the figures,
``same protocol'' means same laboratory, same Hi-C protocol and same
restriction enzyme, and ``different protocol'' means that any of the three
is different.

We also used array-based copy-number segmentation of the two cell lines
obtained from the COSMIC database \citep{forbes2010cosmic} as an external
reference for validation.

All data were processed through a pipeline based on TADbit
\citep{serra2016structural}. Briefly, after controlling the quality of
FASTQ files, paired-end reads were mapped to the corresponding reference
genome (hg38 or mm10) taking into account the restriction enzyme site.
Non-informative contacts were removed applying the following TADbit
filters: \texttt{self-circle}, \texttt{dangling-ends}, \texttt{error},
\texttt{extra-dangling-ends}, \texttt{duplicated} and
\texttt{random-breaks} (for more details see BioRxiv:
https://doi.org/10.1101/132456). In addition, the pipeline is available
from the Supplementary Material published by \citet{quilez2017managing}.

We developed the routines contained in the \texttt{dryhic} R package
(available at http://www.github.com/qenvio/dryhic) to efficiently create
sparse representations of contact matrices and further apply
\textit{vanilla}, \textit{ICE}, \textit{KR} and \textit{oneD} corrections.
\texttt{HiTC} \citep{servant2012hitc} and \texttt{HiCapp}
\citep{wu2016computational} were used to carry out the \textit{LGF} and
\textit{caICB} corrections respectively. When not specified, the
resolution of the analysis is 100~Kbp.

@@

Simulations were based on altering four experimental Hi-C samples obtained
from diploid mouse cell lines (b7fa2d8db\_bfac48760, fc3e8b36a\_7bf1bf374,
GSM987817 and GSM862720, see Supplementary Table~1), that were performed
in either B cells or ES cells, and from either CRG or UCSC. The simulation
strategy was the following: First we selected uniformly at random the
amount of copy number break points (from 3 to 10). We then placed the
break points along chromosomes 18 and 19 uniformly at random. Next, we
assigned a copy number (2, 3, 4 or 10 with equal probability) to each
segment delimited by the break points. We computed the outer product of
this simulated copy number profile and multiplied it element-wise with the
original contact matrices of chromosomes 18 and 19. The resulting matrices
were used as input for correction methods. For each simulation (100 in
total), we measured the pairwise reproducibility score defined by
\citet{yan2017hicspector} and computed the average for pairs from the same
cell type minus the average for pairs from different cell types.

@@

There is no universally accepted standard to compare Hi-C matrices. The
simplest metric is the Spearman correlation applied to intra-chromosomal
contacts up to a given distance (5 Mbp in what follows). The second option
is to measure the similarity of observed over expected contacts via the
Pearson correlation up to a given distance range. Compared to the first,
this metric gives more weight to changes occurring away from the diagonal.
The third option is to compute a correlation per distance stratum and then
obtain a stratum-adjusted correlation coefficient (SCC) as defined by
\citet{yang2017hicrep}. Finally, the ``reproducibility score'' proposed by
\citet{yan2017hicspector} sums the distances between the leading twenty
eigenvectors of the Laplacian of the Hi-C matrix. This approach borrows
the concepts of spectral clustering \citep{von2007tutorial} and amounts to
comparing high level features of the matrix.

We defined five data sets to measure experimental reproducibility after
normalization: The first contained the samples from T47D plus two samples
from K562, the second contained the samples from K562 plus two samples
from T47D, the third contained all the mouse samples, the fourth contained
all the GM12878 samples and the fifth contained one sample of each breast
cancer cells (see Supplementary Table~1 for details). Given a set of
experiments and a metric, we first computed all pairwise combinations
between experiments and then classified the comparisons according to the
characteristics of each pair (cell type, protocol, batch and treatment).

To measure the gain or loss of similarity upon normalization, we compared
raw matrices to obtain a baseline. The differences with this baseline were
estimated using a linear mixed model fitted with the \texttt{lmer}
function of the \texttt{lme4} R package \citep{bates2015lme4}, where the
fixed effect was the normalization method and the random effect was the
chromosome. To test the ability of the different methods to separate
samples from different cell origin we generated receiver operating
characteristic (ROC) curves using the similarity metric (e.g.
reproducibility score) as the classifier score and the relative cell type
(i.e. same \textit{vs} different) as the binary classification. ROC curves
were computed using the \texttt{ROCR} package \citep{sing2005rocr}.
